Inferring the existence and direction of causal associations in the face of measurement error

### 2016-12-29 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Determining the existence and direction of causality in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existence and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{28 December 2016}

@@

With our ability to characterize the human phenome ever improving it is
becoming increasingly common to use statistical tests to infer the
causal relationships between correlated variables. A simple and commonly
used method for inferring if an exposure is on the causal pathway to an
outcome is through using genetic instruments in mediation analysis,
where the effect of an instrument on the outcome is tested before and
after adjusting for the exposure. We show that in the face of
measurement error this can lead to erroneous results, and that
increasing sample size, rather than reducing bias, can often have the
effect of increasing certainty in the wrong answer. Commonly used tools
that are predicated on this methodology such as the causal inference
test (CIT) are shown to be susceptible. We demonstrate that Mendelian
randomization (MR) is a method for causal inference that is robust to
measurement error, and that a simple extension to MR can provide a
method to estimate the direction of causality between correlated
variables even when the biology of the instrument is unknown. Using this
method we demonstrate that for correlations between DNA methylation and
gene expression levels, DNA methylation is likely to be the causal
factor. We show that using MR to infer causal directions is
straightforward, but because measurement error is ubiquitous in
observational data, mediation-based causal inference in particular
should be treated with caution.

@@

The last decade has seem a rapid acceleration in technological advances
that enable us to measure the human phenome in ever increasing detail,
furnishing biologists with the tantalising opportunity to map the
molecular basis of complex traits. This does, however, require
overcoming the statistical challenge of correctly defining the causal
architectures that underlie these high dimensional data. For example,
identifying metabolic perturbations that lead to
disease\textsuperscript{1}, or understanding the influence of epigenomic
influences on gene expression\textsuperscript{2} all depend on
understanding the causal nature of any putative observational
associations. Despite the ubiquity of the question, commonly used
observational study data are notoriously susceptible to pitfalls in
making causal inference, with simple regression-based techniques unable
to distinguish a true causal association from reverse causation or
confounding\textsuperscript{3}.

An established statistical design that can resolve this problem is
randomisation. Given the hypothesis that an exposure is causally related
to an outcome, randomisation can be employed to assess the causal nature
of the association by randomly splitting the sample into two groups,
subjecting one group to the exposure and leaving the other as a control.
The association between the exposure and the outcome in this setting
provides a robust estimate of the causal relationship. This is the
theoretical basis behind randomised control trials, but in practice
randomisation is often impossible to implement in an experimental
context due to cost, scale or inability to manipulate the exposure. The
principle, however, can be employed in extant observational data through
the use of genetic variants associated with the exposure (instruments),
where the inheritance of an allele serves as a random lifetime
allocation of differential exposure levels\textsuperscript{3}. Two
statistical approaches to exploiting the properties of genetic
instruments are widely used: mediation and Mendelian randomisation (MR).

Mediation is a commonly used statistical method for making causal
inference in observational data, and it exists in many forms, from
regression based systems\textsuperscript{4} to structural equation
modeling (SEM)\textsuperscript{5}. In its simplest form mediation
analysis seeks to explain the mechanism that underlies an observed
relationship between an independent and a dependent variable by
introducing a third, `mediator' variable. If the association between the
independent and dependent variables disappears when conditioning on the
mediator variable then this provides evidence that the independent
variable influences the dependent variable through the mediator.

This simple form, however, is prone to being unreliable. Alternative
models of reverse causation or confounding are often indistinguishable
from the hypothesised model of the independent variable causing the
dependent variable via the mediator\textsuperscript{6}. Introducing
pseudo-randomisation by using a genetic instrument as the independent
variable can be used to improve on these issues. Supposing an exposure
has a causal effect on an outcome, then an `instrument' that causes the
exposure (a single nucleotide polymorophism (SNP), for example) should
also influence the outcome. Therefore the influence of the instrument on
the outcome conditional on the exposure should be zero. This forms the
basis of a number of methods such as genetical
genomics\textsuperscript{7}, the regression-based causal inference test
(CIT)\textsuperscript{8,9}, an SEM implementation in the NEO
software\textsuperscript{5}, and various other methods including
Bayesian approaches\textsuperscript{10}. They have been employed by a
number of recent publications that make causal inferences, often in
large scale `omics datasets\textsuperscript{10--15}.

MR employs the same data under but uses a different approach in how the
instrumental variable is employed. Here an instrumental variable
(typically a SNP) that has a robost, direct association with the
exposure is used as a surrogate for the exposure in an association test
with the outcome, avoiding the problems of confounding or reverse
causation by testing for association between the outcome and the SNP.
The causal association between the exposure and the outcome can be
estimated by scaling the association between the SNP and the outcome by
the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation and MR
models have properties that confer some advantages and some
disadvantages. The main advantage of mediation over MR is that the true
underlying biology of the genetic instrument doesn't necessarily need to
be known as long as it associates with the putative exposure. In the CIT
framework (described fully in the Methods) for example, the test
statistic is different if you test for the exposure causing the outcome
or the outcome causing the exposure, allowing the researcher to infer
the direction of causality between two variables by performing the test
in both directions and choosing the model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of
causality the instrument must be known to influence the exposure
directly, associating with the outcome only through the exposure. This
requirement can be problematic because if there exists a putative
association between two variables, with the instrumental SNP being
robustly associated with each, it can be difficult to determine which of
the two variables is subject to the direct effect of the SNP (i.e., for
which of the two variables is the SNP a valid instrument?). By
definition, we expect that if the association is causal then a SNP for
the exposure will be associated with the outcome, so if the researcher
erroneously uses the SNP as an instrument for the outcome then they are
likely to see an apparently robust causal association of outcome on
exposure. Such a situation can arise in many scenarios. Genome wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit\textsuperscript{16}. For
example, the same FTO variant that is associated with body mass index
(BMI) is also associated with C-reactive protein (CRP). It is known that
this is because BMI causes CRP\textsuperscript{17}, but supposing we
only had \emph{a priori} knowledge that FTO associated with CRP, and we
used it to instrument CRP in an MR analysis of CRP against BMI, we would
mistakenly infer that CRP causes BMI. Another situation is where the
causal directions between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{18}?

MR does however have some important advantages over mediation based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, using
mediation requires that the exposure, outcome and instrument variables
are all measured in the same data, whereas recent extensions to MR
circumvent this requirement, allowing causal inference to be drawn when
exposure variables and outcome variables are measured in different
samples\textsuperscript{19}. This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes.
Third, and though this is often underappreciated in fields outside
classical epidemiology it is of crucial importance, MR is substantially
more robust to there being measurement error in the exposure and outcome
variables. Indeed instrumental variable (IV) analysis was in part
initially introduced as a correction for measurement
error\textsuperscript{20}, whereas it has been noted that both basic
mediation\textsuperscript{21--24} and mediation methods that use
instrumental variables\textsuperscript{25,26} are prone to be unreliable
in its presence. We argue that this is a serious limitation.

Measurement (or observational) error is the difference between the
measured value of a quantity and its true value. Such variability can
arise through a whole plethora of mechanisms, which are often unique to
the study design and difficult to avoid\textsuperscript{27,28}. Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but they come with the problem of having imperfect resolution,
for instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{29,30}. Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{31}.

Measurement error will of course arise in more low-tech data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{32}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{28}, and its impact has been explored in the
context of mediation analysis in the epidemiologyical literature
extensively\textsuperscript{6,21,24}.

Given the near-ubiquitous presence of measurement error in biological
data it is vital to understand its impact on the tools we use for causal
inference. Here we show using theory and simulations how
non-differential measurement error and measurement imprecision can lead
to unreliable causal inference in the mediation-based CIT method. We
also present an extension to MR that allows researchers to ascertain the
causal direction even when the biology of the instrument is not fully
understood. This improves the utility of MR in cases where mediation
based methods might have otherwise been used preferentially, and we
apply this new method to infer the direction of causation between DNA
methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{33} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{8} is implemented in the R package
\emph{R/cit}\textsuperscript{9}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{34}.

The direction of the causal association was inferred as follows.
Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
of the variables through the other variable) it is desirable to know
which of the variables, \(x\) or \(y\), it has a direct influence on.
This can be achieved by assessing which of the two variables has the
biggest absolute correlation with \(g\) (Appendix 2). This test can be
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{35}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{19} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (or from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{36}.

@@

Two sample MR\textsuperscript{19} was performed using the summary
statistics presented in\textsuperscript{18} for each of 4122
associations between DNA methylation and gene expression levels. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels. To do this we obtained a
list of 458 gene expression - DNA methylation associations as reported
in Shakhbazov et al\textsuperscript{18}. These were filtered to be
located on the same chromosome, have robost correlations after
correcting for multiple testing, and to share a SNP that had a robust
cis-acting effect on both the DNA methylation probe and the gene
expression probe. Using the reported summary data we used the Steiger
test of two independent correlations to identify which direction of
causality was most likely for each association. We found that the causal
direction commonly goes in both directions (Table 1) but methylation
levels were more likely to influence expression levels than vice versa
(probability of expression causing methylation = 0.0163873 (95\% CI
0.3738727 - 0.4874269) amongst NA associations with
\(p_{Steiger} < 0.05\)).

We performed two sample MR\textsuperscript{19} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
always in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{18}, and that there was a
moderate correlation between the magnitude of the causal correlation and
the observational Pearson correlation (Table 1).

@@

In the epidemiological literature issues of measurement error in
mediation is relatively well explored\textsuperscript{6}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though seemingly
unintuitive, increasing sample size does not resolve the issue, rather
it leads to more extreme p-values for the model that predicts the wrong
direction of causality.

@@

Mediation based network approaches are very well
established\textsuperscript{26} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example applying MR to
genetical genomics\textsuperscript{37}, multivariate
MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
through MR\textsuperscript{40} may offer more robust alternatives for
these more complicated problems.

In our simulations we focused on the simple case of a single instrument
in a single sample setting, and assumed that pleiotropy (the influence
of the instrument on the outcome through a mechanism other than the
exposure) was not present. Recent method developments in
MR\textsuperscript{41} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

@@

\hypertarget{ref-Baron1986}{}
4. Baron, R. M. \& Kenny, D. A. The moderator-mediator variable
distinction in social psychological research: conceptual, strategic, and
statistical considerations. \emph{Journal of personality and social
psychology} \textbf{51,} 1173--82 (1986).

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Inferring the existence and direction of causal associations in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{29 December 2016}

@@

With our ability to characterize the human phenome ever improving it is
becoming increasingly common to use statistical tests to infer the
causal relationships between correlated variables. A simple and commonly
used method for inferring if an exposure is on the causal pathway to an
outcome is through using genetic instruments in mediation analysis,
where the effect of an instrument on the outcome is tested before and
after adjusting for the exposure. We show that in the face of
measurement error this can lead to erroneous inference of both existence
of causality and the causal direction, and that increasing sample size,
rather than reducing bias, can often have the effect of increasing
certainty in the wrong answer. Commonly used tools that are predicated
on this methodology such as the causal inference test (CIT) are shown to
be susceptible. We demonstrate that Mendelian randomization (MR) is a
method for causal inference that is robust to measurement error, and
introduce a simple extension to estimate the direction of causality
between correlated variables even when the biology of the instrument is
unknown. This extension is less susceptible to measurement error than
mediation-based analyses. Using these methods we demonstrate empirically
that the inferred direction of causality between correleted DNA
methylation and gene expression levels is highly susceptible to
differences in measurement error between the two assaying platforms,
demanding caution when making such claims about causality.

The last decade has seem a rapid acceleration in technological advances
that enable us to assess the human phenome in ever increasing detail,
furnishing biologists with the tantalising opportunity to map the
molecular basis of complex traits. This does, however, require
overcoming the statistical challenge of correctly defining the causal
architectures that underlie these high dimensional data. For example,
identifying metabolic perturbations that lead to
disease\textsuperscript{1}, or understanding the influence of epigenomic
influences on gene expression\textsuperscript{2} all depend on
understanding the causal nature of any putative observational
associations.

@@

Observational study data are now widely available but notoriously
susceptible to pitfalls in making causal inference, with basic
regression-based techniques unable to distinguish a true causal
association from reverse causation or confounding\textsuperscript{3--5}.
In response to this, genetic instrumentation has emerged as a valuable
technique for improving the reliability of causal inference in
observational data, and with the coincident rise in genome-wide
association studies it is now a prominent tool that is applied in
several different guises, and has been applied many forms of high
throughput 'omic variables. However, an often neglected pitfall to these
approaches is the influence of non-differential measurement error on the
reliability of causal inference.

Non-differential measurement error (or observational error, henceforth
referred to simply as measurement error) is the difference between the
measured value of a quantity and its true value. Such variability can
arise through a whole plethora of mechanisms, which are often unique to
the study design and difficult to avoid\textsuperscript{6,7}. Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but they come with the problem of having imperfect resolution,
for instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{8,9}. Relatedly, if the measurement of biological
interest is the methylation level in a T cell, then measurement error of
this value can be introduced by using methylation levels from whole
blood samples because the measured value will be an assay of many cell
types\textsuperscript{10}.

Measurement error will of course arise in more low-tech data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{11}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{7}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{12--14}.

Given the near-ubiquitous presence of measurement error in phenomic data
it is vital to understand its impact on the tools we use for causal
inference. An established study design that can provide information
about causality is randomisation. Given the hypothesis that trait A
(henceforth referred to as the exposure) is causally related to trait B
(henceforth referred to as the outcome), randomisation can be employed
to assess the causal nature of the association by randomly splitting the
sample into two groups, subjecting one group to the exposure and leaving
the other as a control. The association between the exposure and the
outcome in this setting provides a robust estimate of the causal
relationship. This provides the theoretical basis behind randomised
control trials, but in practice randomisation is often impossible to
implement in an experimental context due to cost, scale or inability to
manipulate the exposure. The principle, however, can be employed in
extant observational data through the use of genetic variants associated
with the exposure (instruments), where the inheritance of an allele
serves as a random lifetime allocation of differential exposure
levels\textsuperscript{3}. Two statistical approaches to exploiting the
properties of genetic instruments are widely used: mediation-based
approaches and Mendelian randomisation (MR).

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical
genomics\textsuperscript{15}, the regression-based causal inference test
(CIT)\textsuperscript{16,17}, a structural equation modelling (SEM)
implementation in the NEO software\textsuperscript{18}, and various
other methods including Bayesian approaches\textsuperscript{19}. They
have been employed by a number of recent publications that make causal
inferences in large scale `omics datasets\textsuperscript{19--24}.

MR can operate on the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP, which is assumed to have a direct association with the
exposure, is used as a surrogate for the exposure. The causal effect of
the exposure and the outcome can be estimated by scaling the association
between the SNP and the outcome by the association between the SNP and
the exposure.

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. The main advantage of
mediation over MR is that the true underlying biology of the genetic
instrument doesn't necessarily need to be known as long as it associates
with the putative exposure. In the CIT framework (described fully in the
Methods) for example, the test statistic is different if you test for
the exposure causing the outcome or the outcome causing the exposure,
allowing the researcher to infer the direction of causality between two
variables by performing the test in both directions and choosing the
model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of
causality the instrument must be known to influence the exposure
directly, associating with the outcome only through the exposure. This
requirement can be problematic because if there exists a putative
association between two variables, with the instrumental SNP being
robustly associated with each, it can be difficult to determine which of
the two variables is subject to the direct effect of the SNP (i.e., for
which of the two variables is the SNP a valid instrument? Figure 1). By
definition, we expect that if the association is causal then a SNP for
the exposure will be associated with the outcome, so if the researcher
erroneously uses the SNP as an instrument for the outcome then they are
likely to see an apparently robust causal association of outcome on
exposure. Such a situation can arise in many scenarios. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit\textsuperscript{25}. For
example, the same FTO variant that is associated with body mass index
(BMI) is also associated with C-reactive protein (CRP). It is known that
this is because BMI causes CRP\textsuperscript{26}, but supposing we
only had \emph{a priori} knowledge that FTO associated with CRP, and we
used it to instrument CRP in an MR analysis of CRP against BMI, we would
mistakenly infer that CRP causes BMI. Another situation is where the
causal directions between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{27}
(Figure 1)?

MR does however have some important advantages over mediation based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, using
mediation requires that the exposure, outcome and instrument variables
are all measured in the same data, whereas recent extensions to MR
circumvent this requirement, allowing causal inference to be drawn when
exposure variables and outcome variables are measured in different
samples\textsuperscript{28}. This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated. Third, MR is substantially more robust to there
being measurement error in the exposure and outcome variables. Indeed
instrumental variable (IV) analysis was in part initially introduced as
a correction for measurement error\textsuperscript{29}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{13,14,30,31} and mediation-based methods that
use instrumental variables\textsuperscript{32,33} are prone to be
unreliable in its presence. We argue that this is a serious limitation.

Here we show using theory and simulations how non-differential
measurement error and measurement imprecision can lead to unreliable
causal inference in the mediation-based CIT method. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and to evaluate the sensitivity of the result to
measurement error. This improves the utility of MR in cases where
mediation based methods might have otherwise been used preferentially.
Finally, we apply this new method to infer the direction of causation
between DNA methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{34} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{16} is implemented in the R package
\emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{35}.

The direction of the causal association was inferred as follows.
Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
of the variables through the other variable) it is desirable to know
which of the variables, \(x\) or \(y\), it has a direct influence on.
This can be achieved by assessing which of the two variables has the
biggest absolute correlation with \(g\) (Appendix 2). This test can be
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{36}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (or from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{37}.

@@

Two sample MR\textsuperscript{28} was performed using the summary
statistics presented in\textsuperscript{27} for each of 4122
associations between DNA methylation and gene expression levels. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels. To do this we obtained a
list of 458 gene expression - DNA methylation associations as reported
in Shakhbazov et al\textsuperscript{27}. These were filtered to be
located on the same chromosome, have robost correlations after
correcting for multiple testing, and to share a SNP that had a robust
cis-acting effect on both the DNA methylation probe and the gene
expression probe. Using the reported summary data we used the Steiger
test of two independent correlations to identify which direction of
causality was most likely for each association. We found that the causal
direction commonly goes in both directions (Table 1) but methylation
levels were more likely to influence expression levels than vice versa
(probability of expression causing methylation = 0.0163873 (95\% CI
0.3738727 - 0.4874269) amongst NA associations with
\(p_{Steiger} < 0.05\)).

We performed two sample MR\textsuperscript{28} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
always in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{27}, and that there was a
moderate correlation between the magnitude of the causal correlation and
the observational Pearson correlation (Table 1).

@@

In the epidemiological literature issues of measurement error in
mediation is relatively well explored\textsuperscript{12}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though seemingly
unintuitive, increasing sample size does not resolve the issue, rather
it leads to more extreme p-values for the model that predicts the wrong
direction of causality.

@@

Mediation based network approaches are very well
established\textsuperscript{33} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example applying MR to
genetical genomics\textsuperscript{38}, multivariate
MR\textsuperscript{39}, network MR\textsuperscript{40} and mediation
through MR\textsuperscript{41} may offer more robust alternatives for
these more complicated problems.

In our simulations we focused on the simple case of a single instrument
in a single sample setting, and assumed that pleiotropy (the influence
of the instrument on the outcome through a mechanism other than the
exposure) was not present. Recent method developments in
MR\textsuperscript{42} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

@@

\hypertarget{ref-Phillips1991}{}
4. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
effects? relative risk estimation when correlated exposures are measured
imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
1223--1231 (1991).
